---
title: 'Vignette 7: Calculation of distances between proteins'
author: "DFM & PL"
date: "`r Sys.Date()`"
output:
  #word_document:
  #  toc: yes
  html_document:
    toc: yes
    df_print: paged
  #pdf_document:
  #  toc: yes
  #  df_print: paged
vignette: |
  %\VignetteIndexEntry{Vignette 7: Calculation of distances between proteins}     
  %\VignetteEngine{knitr::rmarkdown}
---
 
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "#>",
  fig.width = 4,
  fig.height = 4,
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(
    keep.blank.line = TRUE,
    width.cutoff = 150
  ),
  options(width = 150),
  eval = TRUE
)
```

## Finding proteins with similar profiles

We may find the proteins with profiles nearest to a given protein using the function "nearestProts". Distance is computed as the Euclidean distance between profiles. To use the function, we first use the R function `dist` to create a distance matrix for the proteins in a list of mean profiles, such as `protProfileNSA_AT5tmtMS2`. For clarity of presentation, we rename the embedded data sets to remove experiment-specific labels.

```{r, echo=TRUE, eval=TRUE}
library(protlocassign)
data(protNSA_AT5tmtMS2)
data(totProtAT5)
protNSA <- protNSA_AT5tmtMS2
totProt <- totProtAT5
distUseNSA <- dist(protNSA[,1:9], method="euclidean")
```

Then select the protein names:

```{r, echo=TRUE, eval=TRUE}
protsUse <- rownames(protNSA)
```
Finally, provide a protein name. Here, for the protein "CTSD", we find the 10 nearest proteins.

```{r, echo=TRUE, eval=TRUE}
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUseNSA, protNames=protsUse,
             profile=protNSA)
```

Instead of using normalized specific amounts, we may transform them to relative specific amounts:

```{r, echo=TRUE, eval=TRUE}
protProfileLevelsRSA <- RSAfromNSA(NSA=protNSA[,1:9],
                                 NstartMaterialFractions=6, totProt=totProt)
distUseRSA <- dist(protProfileLevelsRSA, method="euclidean")
nearestProts(protName="CTSD", n.nearest=10,  distProts=distUseRSA, protNames=protsUse,
             profile=protProfileLevelsRSA)

```

Note that if one wants to generate a table listing the distances between all protein pairs, one needs to convert the distUse or distUseRSA to a matrix. We show the first five rows and columns here: 

```{r, echo=TRUE, eval=TRUE}
distUseNSAmatrix <- as.matrix(distUseNSA)
distUseNSAmatrix[1:5,1:5]
```

This matrix can be written to a local directory using standard procedures.

## Reproducibility

```{r, echo=TRUE, eval=TRUE}
utils::sessionInfo()
```


